---
title: "Mental Health Treatment Prediction in the Workplace"
subtitle: "Preparation 1: Data Selection and Initial Preprocessing"
author: "Richel Attafuah"
date: "April 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
```

```{r libraries, message=FALSE, warning=FALSE,include=FALSE}
library(tidyverse)
library(kableExtra)
library(scales)
```

## 1. Data Overview & Source

The dataset used in this project is the **OSMI Mental Health in Tech Survey (2014)**, collected by Open Sourcing Mental Illness (OSMI), a non-profit organization dedicated to raising awareness and providing resources for mental health in the tech community. The survey was administered online and captures responses from tech workers across 60+ countries on workplace mental health culture, employer support, and personal 
mental health history.

**Source:** [OSMI Mental Health in Tech Survey — Kaggle](https://www.kaggle.com/datasets/osmi/mental-health-in-tech-survey)


```{r load-data, message=FALSE,warning=FALSE, include=FALSE}
survey <- read_csv("data/survey.csv")
dim(survey)
glimpse(survey)
```

## 2. Response Variable Identification

The response variable for this project is **`treatment`** ,  a binary variable indicating whether a tech worker has sought mental health treatment (Yes/No). This is a **Classification** problem. The goal is to predict the probability that a 
tech worker seeks mental health treatment based on workplace environment, employee support policies, and personal background factors.

```{r response-var}
survey %>%
  count(treatment) %>%
  mutate(proportion = round(n / sum(n), 3)) %>%
  kbl(caption = "Response variable distribution",
      col.names = c("Sought Treatment", "Count", "Proportion")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#2E86AB", color = "white")
```

## 3. Preliminary Exploratory Data Analysis (EDA)

### 3.1 Dataset Structure

The dataset contains **1,259 respondents** and **27 variables**. After removing non-informative columns (`Timestamp`, `state`, `comments`), we retain **24 predictors**.The dataset comfortably exceeds the 8–10 predictor requirement.


```{r eda-structure}
survey %>%
  summarise(
    total_variables = ncol(.),
    numeric_vars    = sum(sapply(., is.numeric)),
    character_vars  = sum(sapply(., is.character)),
    datetime_vars   = sum(sapply(., lubridate::is.timepoint))
  ) %>%
  kbl(caption = "Dataset structure summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

### 3.2 Variable Types

```{r variable-types}
tibble(
  Variable = names(survey),
  Type     = sapply(survey, class),
  Role     = case_when(
    Variable == "treatment"  ~ "Response",
    Variable %in% c("Timestamp", "state", "comments") ~ "Drop",
    TRUE ~ "Predictor"
  )
)
```

### 3.3 Response Variable Distribution

```{r response-plot, fig.align='center'}
survey %>%
  count(treatment) %>%
  mutate(proportion = round(n / sum(n), 3)) %>%
  ggplot(aes(x = treatment, y = n, fill = treatment)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", proportion*100, "%)")),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Yes" = "#2E86AB", "No" = "#A8DADC")) +
  labs(title    = "Distribution of mental health treatment-seeking",
       subtitle = "Nearly balanced classes — no resampling correction needed",
       x        = "Sought treatment",
       y        = "Count") +
  theme_minimal(base_size = 13)
```

### 3.4 Missing Data Overview

```{r missing-data}
survey %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  filter(Missing > 0) %>%
  arrange(desc(Missing)) %>%
  mutate(Pct_Missing = round(Missing / nrow(survey) * 100, 1)) %>%
  kbl(caption = "Missing data by variable") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#2E86AB", color = "white")
```

## 4. Initial Preprocessing & Cleaning

### 4.1 Drop non-informative variables

```{r drop-cols,include=FALSE}
survey_clean <- survey %>%
  select(-Timestamp, -state, -comments)

dim(survey_clean)
```

### 4.2 Clean the Gender variable

```{r clean-gender}
survey_clean <- survey_clean %>%
  mutate(Gender = case_when(
    str_to_lower(Gender) %in% c("male", "m", "man", "cis male",
                                 "male (cis)", "maile", "mal",
                                 "make", "male ", "msle") ~ "Male",
    str_to_lower(Gender) %in% c("female", "f", "woman", "cis female",
                                 "female (cis)", "femail", "femake",
                                 "female ") ~ "Female",
    TRUE ~ "Other"
  ))

survey_clean %>%
  count(Gender) %>%
  kbl(caption = "Gender distribution after cleaning",
      col.names = c("Gender", "Count")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#2E86AB", color = "white")
```

### 4.3 Fix Age outliers

```{r clean-age,include=FALSE}
# Inspect Age range before cleaning
summary(survey_clean$Age)

# Remove implausible ages (keep 18-75)
survey_clean <- survey_clean %>%
  filter(Age >= 18, Age <= 75)

summary(survey_clean$Age)
```

### 4.4 Impute missing values

```{r impute-missing,include=FALSE}
# Impute work_interfere and self_employed with mode
get_mode <- function(x) {
  ux <- unique(na.omit(x))
  ux[which.max(tabulate(match(x, ux)))]
}

survey_clean <- survey_clean %>%
  mutate(
    work_interfere = ifelse(is.na(work_interfere), 
                            get_mode(work_interfere), work_interfere),
    self_employed  = ifelse(is.na(self_employed),  
                            get_mode(self_employed),  self_employed)
  )

# Confirm no missing values remain
colSums(is.na(survey_clean))
```

### 4.5 Convert characters to factors

```{r factorize,include=FALSE}
survey_clean <- survey_clean %>%
  mutate(across(where(is.character), as.factor))

glimpse(survey_clean)
```

### 4.6 Final cleaned dataset summary


```{r final-summary}
dim(survey_clean)

survey_clean %>%
  summarise(
    n_observations = n(),
    n_predictors   = ncol(.) - 1,
    numeric_vars   = sum(sapply(., is.numeric)),
    factor_vars    = sum(sapply(., is.factor))
  ) %>%
  kbl(caption = "Final cleaned dataset summary",
      col.names = c("Observations","Predictors","Numeric vars","Factor vars")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#2E86AB", color = "white")
```